Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-23 manuscript.qmd

## Consensus on Involvement Status

@@

These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and patterns of involvement. In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by T-category, ipsilateral number of involved LNLs, and whether the tumor extended over the mid-sagittal line. A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.

@@

The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated. For example, midline extension was present in 48% of advanced T-category tumors but only in 7% of early T-category tumors. The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors. However, @fig-data-strat-uncorr indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors . The figure considers only patients with lateralized tumors not extending over the midline. For early T-category and healthy ipsilateral levels I-V, only 1% (1 out of 84 patients) shows involvement of contralateral level II. This increases to 5% (9 out of 165 patients) if ipsilateral level II is involved, to 15% (9 out of 60 patients) if ipsilateral levels II and III are involved, and to 23% (6 out of 26 patients) for advanced T-category with ipsilateral levels II and III involved.<!-- TODO: make values here dynamic. -->

@@

Based on the observations of the [previous section](#sec-data-strat), any potential model that aims to also predict the risk for contralateral nodal involvement, should be able to take the following into account:

@@

Over the course of this work, we will first briefly recap the mentioned HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only. Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra. In this section, we also introduce a way of modelling the tumor's midline extension as a random variable (@sec-midline) and lastly talk about how it may affect the contralateral spread in @sec-params-symmetry.

@@

## Likelihood Function

@@

<!--There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->

@@

In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters. I.e., the ipsi- and contralateral spread rates are entirely different. However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the sides:

@@

1. The tumor was still lateralized at $t=\tau$ and just grew over the midline. Under this scenario, the probability for having a midline extension at $t=\tau+1$ is given by $p_\epsilon$. We use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
2. The mid-sagittal line was already crossed by the tumor. In that case, the probability is 1 to remain in that lateralization state. Thus, for this case we simply add the same distribution we are computing, but from one time-step earlier, arriving at a recursion relation:

@@

It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity. However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference. We opted for this because the literature values of sensitivity and specificity [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging. This is virtually impossible with sensitivities around 80% and specificities lower than 100%.

Our pre-training consensus was formed by considering all reported diagnostic information for a particular patient and LNL. When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].

>>>>

## Consensus on Involvement Status {#sec-data-consensus}

@@

These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement. In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line. A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.

@@

The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated. For example, midline extension was present in 48% of advanced T-category tumors but only in 7% of early T-category tumors. The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors. However, @fig-data-strat-uncorr indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors. The figure considers only patients with lateralized tumors not extending over the midline. For early T-category and healthy ipsilateral levels I-V, only 1% (1 out of 84 patients) shows involvement of contralateral level II. This increases to 5% (9 out of 165 patients) if ipsilateral level II is involved, to 15% (9 out of 60 patients) if ipsilateral levels II and III are involved, and to 23% (6 out of 26 patients) for advanced T-category with ipsilateral levels II and III involved.<!-- TODO: make values here dynamic. -->

@@

Based on the observations of the [previous section](#sec-data-strat), any model to predict the risk for contralateral nodal involvement, should be able to describe the following:

@@

Over the course of this work, we will first briefly recap the HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only. Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra. In this section, we also introduce a way of modelling the tumor's midline extension as a random variable (@sec-midline) and lastly talk about how it may affect the contralateral spread in @sec-params-symmetry.

@@

## Likelihood Function of the Unilateral Model

@@

<!-- There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->

@@

In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters, i.e., the ipsi- and contralateral spread rates are entirely different. However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the two sides:

@@

1. The tumor was still lateralized at $t=\tau$ and just grew over the midline. Under this scenario, the probability for having a midline extension at $t=\tau+1$ is given by $p_\epsilon$. We use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
2. The mid-sagittal line was already crossed by the tumor. In that case, the probability is 1 to remain in that lateralization state. Thus, to account for this case we simply add the distribution $P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)$ from one time-step earlier, arriving at a recursion relation:

@@

It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity. However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference, as introduced in @sec-sec-data-consensus. We opted for this because the literature values of sensitivity and specificity [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging. This is virtually impossible with sensitivities around 80% and specificities lower than 100%.
